---
title: "Introduction to libinsight"
author: "Sahl Mohammed"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to libinsight}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%"
)
```

## Overview

libinsight is an R package for the analysis and visualization of academic
library circulation statistics. For library professionals crafting an annual
report, researchers examining information behavior, and LIS students interested
in data analysis, libinsight provides a comprehensive suite of utilities to
derive insights from raw circulation data.

With libinsight, you will find:

- A specific S3 object (`libdata`) for organizing library data sets, complete
  with specialized `print`, `summary`, and `plot` methods.
- Seven analytical functions for summarizing circulation data, generating
  trends, ranking top circulation items, characterizing patrons, evaluating
  collection age, creating circulation heatmaps, and automatically generating
  text-based reports.
- A built-in sample dataset (`circ_sample`) enabling immediate data exploration
  without requiring your own data source.

---

## Installation

Install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("sahlmohammedwork-commits/libinsight")
```

Then load the package:

```{r load-pkg}
library(libinsight)
```

---

## The S3 class: `libdata`

All functions in `libinsight` accept either a plain `data.frame` or a
`libdata` object. Using `libdata` is recommended because the constructor
validates your data up front, and the S3 methods give you instant overviews.

### Required columns

| Column | Type | Description |
|---|---|---|
| `item_id` | character | Unique item identifier |
| `title` | character | Item title |
| `subject` | character/factor | Subject category |
| `pub_year` | integer | Publication year |
| `patron_type` | character/factor | Patron category |
| `checkout_month` | integer (1–12) | Month of checkout |
| `checkout_year` | integer | Year of checkout |
| `checkouts` | integer | Checkout count for this row |

### Creating a `libdata` object

```{r create-libdata}
data(circ_sample)        # load the built-in sample data
ld <- libdata(circ_sample)
class(ld)
```

### `print` method

```{r print-libdata}
print(ld)
```

### `summary` method

```{r summary-libdata}
s <- summary(ld)
```

### `plot` method

The default plot shows total checkouts by subject category.

```{r plot-libdata}
plot(ld)
```

---

## Analyzing circulation statistics

### `summarize_circulation()`

Compute total, mean, median, min, max checkouts and identify peak periods.

```{r summarize}
summarize_circulation(ld)
```

Filter to a specific year range:

```{r summarize-range}
summarize_circulation(ld, year_from = 2020, year_to = 2022)
```

---

### `plot_circ_trend()`

Visualize how checkout volume changes over time.

```{r trend-monthly}
plot_circ_trend(ld)
```

Switch to annual aggregation:

```{r trend-annual}
plot_circ_trend(ld, by = "year")
```

---

### `top_items()`

Identify the most circulated titles.

```{r top-items-all}
top_items(ld, n = 5)
```

Filter by subject or patron type:

```{r top-items-filtered}
top_items(ld, n = 5, subject = "Technology")
top_items(ld, n = 5, patron_type = "Graduate")
```

---

### `patron_stats()`

Break down usage by who is checking items out.

```{r patron-stats}
ps <- patron_stats(ld)
```

---

### `collection_age()`

Examine whether the most-used materials are current or aging. The histogram
is weighted by checkout count so high-circulation titles have more visual weight.

```{r collection-age}
collection_age(ld)
```

---

### `lib_heatmap()`

A tile heatmap reveals seasonal patterns across patron types simultaneously.

```{r heatmap}
lib_heatmap(ld)
```

---

### `lib_report()`

Generate a comprehensive plain-text report combining all the analyses above.

```{r lib-report}
lib_report(ld, title = "Academic Library Annual Report", top_n = 5)
```

---

## Workflow example: uploading your own data

If you have real circulation data in a CSV file, the typical workflow is:

```r
# 1. Read your data
my_data <- read.csv("my_library_data.csv", stringsAsFactors = FALSE)

# 2. Ensure required column names match
# (rename if necessary)
names(my_data)

# 3. Wrap in a libdata object
ld_real <- libdata(my_data)

# 4. Run any analysis
lib_report(ld_real, title = "My Library Report 2025")
plot_circ_trend(ld_real, by = "year")
top_items(ld_real, n = 10)
```

---

## Package metadata

| Field | Value |
|---|---|
| Package | libinsight |
| Version | 1.0.0 |
| License | GPL-2 |
| Author | Sahl Mohammed |
| Repository | https://github.com/sahlmohammedwork-commits/libinsight |
| Depends | R (>= 4.0.0) |
| Imports | ggplot2, dplyr |

---

## Session information

```{r session-info}
sessionInfo()
```
